---
title: "Burden LE"
author: "Ivan Hanigan, Richard Broome and Josh Horsley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{burden_le}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

library(iomlifetR)
```

# The life tables 'Chiang method (II)' for subnational life expectancy at birth results

This vignette demonstrates years/days of life gained for a reduction in PM2.5 with 95% confidence limits.

It uses the 'Chiang method (II)' for Life Table calculations (ref Chiang CL. The Life Table and its Applications.Malabar (FL): Robert E Krieger Publ Co, 1984).
 
Data source: Life Expectancy at Birth and at Age 65 by Local Areas in the United Kingdom: 2004?06 to 2008?10 Office for National Statistics (ONS)
http://www.ons.gov.uk/ons/rel/subnational-health4/life-expec-at-birth-age-65/2004-06-to-2008-10/ref-life-table-template.xls.

For this example we will use the HRAPIE mean relative risk estimate RR = 1.062 (1.040-1.083) per 10 ?g/m3 of PM2.5. (Ref Health risks of air pollution in Europe ?HRAPIE project Recommendations for concentration?response functions for cost-benefit analysis of particulate matter, ozone and nitrogen dioxide World Health Organization 2013 http://www.euro.who.int/__data/assets/pdf_file/0006/238956/Health_risks_air_pollution_HRAPIE_project.pdf?Ua=1)

```{r}
## Scenario A: current exposure level of PM2.5
## Scenario B: reduction of 3ug/m3 of PM2.5

##*****************************************************

##Install Libraries

library(data.table)
library(iomlifetR)


## Using an abridged set of age, population and mortality data for 5 year age groups provided
dat <- chiang_demo

## This package uses demographic data in a standardised format of age, population, deaths stored in demog_data. 
## Create demog data
demog_data = data.frame(age = dat$start_age,
                            population = dat$pop,
                            deaths = dat$death)
demog_data

## Set up scenario:
## use HRAPIE recommended concentration?responsefunctions (CRFs)given as relative risks (Table 1 Page 11)
## for PM2.5  
rr_use <- data.frame(est_type = c("RRper10", "RRper10_lci", "RRper10_uci"),
                     est = c(1.062, 1.040, 1.083))
rr_use

## estimate for change in exposure of PM2.5, set to 3ug/m3

deltaX = 3
#deltaX = 5

## Calculate for RR and 95CIs (using loop)


for(est_type in rr_use$est_type){
    ##
    rr <- rr_use[rr_use$est_type == est_type, "est"]
    print(est_type)
    
## Calculate life table for baseline (Scenario A)and reduced exposure (Scenario B)scenarions
##
le = burden_le(demog_data,
               pm_concentration = deltaX,
               RR = rr)
 le

## Calculate years of life lost yll from lifetable

impacted_le <- le[["impacted"]][, "ex"]
 impacted_le

an <- burden_an(demog_data,
                pm_concentration = deltaX,
                RR = rr)
yll <- burden_yll(an, impacted_le)
yll
## results in a data frame 'burden' 
burden <- data.frame(x_attrib = an, population = demog_data$population, le, an, yll)
    ##
burden$est_type <- est_type

if(est_type == "RRper10"){
    burden_out <- burden
} else {
    burden_out <- rbind(burden_out, burden)
}
}
## end of loop


##burden_out
setDT(burden_out)    
burden_out[,sum(yll), by = est_type]

## Prints years of life lost (V1) with 95% confidence limits for scenario A (current PM2.5 level)
## compared to a 3ug/m3 reduction (Scenario B) 
"
      est_type       V1
1:     RRper10 566.8479
2: RRper10_lci 369.4309
3: RRper10_uci 751.6683
"
##burden_out[1,]
ledif <- data.frame(
    impacted_le = burden_out[baseline.age == 0, impacted.ex],
    baseline_le = burden_out[baseline.age == 0, baseline.ex]
    )
(ledif[,1] - ledif[,2]) * 365
## Prints days of life gained by birth cohort in scenario B (reduction in 3ug/m3)
"
71.12587
lower 46.33968
upper 94.34513
"
## NOTE: For a reduction of 5ug/m3
## the years of life lost is 
"      est_type        V1
1:     RRper10  945.5301
2: RRper10_lci  616.0461
3: RRper10_uci 1254.1772

and days of life gained by the birth cohort
 118.71546 (7.30533-157.54756)
"
```