---
title: "Impact"
author: "Josh Horsley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Impact}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

library(iomlifetR)
```

This vignette demonstrates impact assessment with various lag choices

# Reduce PM~2.5~ by 1 $\mu g/m^3$

For this example we will use the HRAPIE mean relative risk estiamte RR = 1.06 per 10 $\mu g/m^3$ of PM~2.5~.

## No cessation lag

Functions in this package use demographic data in a standardised format stored in `demog_data`. For this example we will use the included `single_year_data` to create `demog_data`.
```{r}
head(single_year_data)
population <- subset(single_year_data,
                     time == 2011 & sex == "Persons" & measure == "Population")
population <- population[, c("age", "value")]
population$age <- as.numeric(gsub(" .+", "", population$age))
colnames(population)[2] <- "population"
deaths <- subset(single_year_data,
                 time == 2011 & sex == "Persons"& measure == "Deaths")
deaths <- deaths[, "value"]
demog_data <- data.frame(population, deaths = deaths)
```

## Calculate the impact assuming there is no lag between exposure reduction and health outcomes
```{r}
no_lag <- impact(demog_data)
no_lag <- data.frame(lapply(no_lag, colSums))
```

Compare deaths avoided
```{r}
plot(rownames(no_lag), no_lag$deaths_current,
     xlab = "Year", ylab = "Number",
     main = "Deaths avoided", type = "b")
points(rownames(no_lag), no_lag$deaths_ext, col = "red")
abline(h = 0)
legend("topright", c("Current cohort", "Extended population"),
       col=c("black", "red"), pch = 21, lty=2)
```

Comapre life-years gained
```{r}
plot(rownames(no_lag), no_lag$ly_current,
     xlab = "Year", ylab = "Number",
     main = "Life-years gained", type = "b")
points(rownames(no_lag), no_lag$ly_ext, col = "red")
abline(h = 0)
legend("bottom", c("Current cohort", "Extended population"),
       col=c("black", "red"), pch = 21, lty=2)
```



## Calculate the impact assuming the US EPA lag between exposure reduction and health outcomes

```{r}
lag <- cumsum(c(0.3, rep(0.5/4, 4), rep(0.2/15, 15)))
epa_lag <- impact(demog_data, lag_structure = lag)
epa_lag <- data.frame(lapply(epa_lag, colSums))
```

Plot a comparison of no lag and US EPA lag (Extended cohort)
```{r}
plot(rownames(no_lag), no_lag$deaths_ext,
     xlab = "Year", ylab = "Number",
     main = "Deaths avoided", type = "b")
points(rownames(epa_lag), epa_lag$deaths_ext, col = "red", type = "b")
abline(h = 0)
legend("topright", c("No lag", "US EPA lag"),
       col=c("black", "red"), pch = 21, lty=2)


```

Compare total life-years gained
```{r}
plot(rownames(no_lag), no_lag$ly_ext,
     xlab = "Year", ylab = "Number",
     main = "Life-years gained" , type = "b")
points(rownames(epa_lag), epa_lag$ly_ext, col = "red", type = "b")
abline(h = 0)
legend("bottomright", c("No lag", "US EPA lag"),
       col=c("black", "red"), pch = 21, lty=2)
```


## Assuming PM~2.5~ takes 10 years to fall by 1 $\mu g/m^3$ and US EPA cessation lag

Calculate results
```{r}
pm <- seq(0.1, 1, 0.1)
slow_pm <- impact(demog_data, delta_pm = pm, lag_structure = lag)
slow_pm <- data.frame(lapply(slow_pm, colSums))

```

Compare deaths avoided
```{r}
plot(rownames(no_lag), no_lag$deaths_ext,
     xlab = "Year", ylab = "Number",
     main = "Deaths avoided", type = "b")
points(rownames(epa_lag), epa_lag$deaths_ext, col = "red", type = "b")
points(rownames(slow_pm), slow_pm$deaths_ext, col = "blue", type = "b")
abline(h = 0)
legend("topright", c("No lag", "US EPA lag", "Lag and gradual fall in PM"),
       col=c("black", "red", "blue"), pch = 21, lty=2)

```

Compare life years gained
```{r}
plot(rownames(no_lag), no_lag$ly_ext,
     xlab = "Year", ylab = "Number",
     main = "Life-years gained", type = "b")
points(rownames(epa_lag), epa_lag$ly_ext, col = "red", type = "b")
points(rownames(slow_pm), slow_pm$ly_ext, col = "blue", type = "b")
abline(h = 0)
legend("bottomright", c("No lag", "US EPA lag", "Lag and gradual fall in PM"),
       col=c("black", "red", "blue"), pch = 21, lty=2)
```

